---
title: "ProteinBatcher workflow"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{ProteinBatcher workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
Suggests:
    knitr,
    rmarkdown,
    htmltools
---

# Overview
Quantitative proteomics is well supported in Bioconductor for individual
downstream tasks such as differential testing (e.g. via limma), missing-value
imputation, or visualization. However, fully integrated downstream workflows
that combine these steps in a single, reproducible, and design-aware pipeline
are comparatively scarce, particularly for label-free proteomics. ProteinBatcher
does not run upstream quantification; it operates on protein-level abundance
matrices (the package description explicitly targets DIA-NN-style outputs).

ProteinBatcher addresses this gap by providing a cohesive downstream analysis
framework that focuses on three aspects that are often difficult to combine in
practice:

1. Flexible experimental designs
    - The package is built around limma and supports:
    - multi-factor designs (e.g. condition + batch, condition + sex),
    - interaction testing, 
    - paired and repeated-measures designs,
    - correlation-aware models via duplicateCorrelation.

    This flexibility allows users to model common proteomics scenarios such as
    batch effects, donor effects, or preparation-day effects without rewriting
    analysis code for each experiment.
    
2. End-to-end standardization of downstream steps. ProteinBatcher integrates,
in a single workflow:
    - missingness-based filtering,
    - mean and left-censored (LDV) imputation strategies,
    - differential abundance testing,
    - systematic separation of main, common, and interaction effects,
    - generation of plot-ready result objects.

    By enforcing a consistent structure for inputs, models, and outputs, the
    package reduces ad hoc analysis choices and improves reproducibility across
    experiments.
    
3. Interpretation-oriented visualizations. In addition to standard volcano
plots, ProteinBatcher implements:
    - deregulograms, which compare full effects across the two levels of an
        interaction factor (e.g. batch day 1 vs day 2, female vs male).
        Deregulograms complement interaction volcano plots by emphasizing effect
        concordance and magnitude, rather than statistical significance alone,
        and facilitate biological interpretation of interaction effects.

Importantly, ProteinBatcher is not intended to replace existing Bioconductor
infrastructure, but rather to orchestrate and standardize well-established
methods (notably limma) into a coherent downstream proteomics workflow.
The package is therefore particularly suited for users who repeatedly analyze
proteomics experiments with similar structure but varying designs, and who need
a balance between automation and modeling flexibility.

This vignette documents the workflow and outputs, but does not execute the
full analysis. All examples rely on precomputed results stored in
inst/extdata, ensuring portability and fast rendering during package checks.

# Relationship with the Bioconductor ecosystem

ProteinBatcher is designed to operate within the standard Bioconductor
infrastructure, rather than introducing new data containers or statistical
methods. Its functionality is built by extending existing Bioconductor
components in a consistent downstream workflow.

The package is built around two Bioconductor pillars:

- **SummarizedExperiment** as the central container for assays, feature metadata
(rowData) and sample metadata (colData).
- **limma** for linear modeling and empirical Bayes moderation.

All intermediate and final results in ProteinBatcher are stored in
SummarizedExperiment objects. Missing-value filtering and imputation are
implemented as modular steps that operate directly on SummarizedExperiment
objects. 

# Input data and workflow parameters

ProteinBatcher needs two input data objects and a set of workflow parameters.
This distinction is important: the inputs define the data to be analysed,
whereas the parameters control how the analysis is performed.

## Input data

ProteinBatcher expects exactly two input datasets:

1. Quantitative matrix: A feature-by-sample matrix containing quantitative
proteomics measurements (typically protein groups). Rows correspond to
quantified features and columns to samples. The matrix is assumed to originate
from an upstream workflow (e.g. FragPipe or DIANN), but ProteinBatcher does not
depend on a specific quantification tool.

2. Sample annotation table: A tab-delimited table describing the experimental
design and linking each sample to its biological and technical metadata.

Both inputs are provided in this vignette as example datasets shipped with the
package.

```{r}
# Load package
library(ProteinBatcher)
library(readr)

# Input files shipped with the package
path_annotation <- system.file(
  "extdata", "annotation_HaCaT.tsv",
  package = "ProteinBatcher"
)

path_pgmatrix <- system.file(
  "extdata",
  "2024MK017_HaCaT_Stimulation_1to24_Astral_report.pg_matrix.tsv",
  package = "ProteinBatcher"
)

# Output directory (portable, not used during vignette build)
path_output <- tempdir()

# Inspect annotation structure
ann <- readr::read_tsv(path_annotation)
```

## Sample annotation structure
The annotation table must contain at least a condition column defining the main
experimental comparison. Additional columns enable more complex designs such as
batch correction, paired analyses, and interaction testing.

The following columns are supported:

- file: raw data file identifier
- sample / sample_name: sample identifiers
- condition: main experimental condition of interest
- replicate: technical replicate identifier (repeated measurements of the
    same sample under the same condition)
- block: biological replicate or donor identifier
- batch: batch or preparation factor (e.g. factor to control for against
    condition)

Only columns explicitly referenced in the design formula are required.

## Workflow parameters

In addition to the input data, ProteinBatcher requires several parameters that
define how the analysis is carried out. These parameters do not represent
additional inputs, but control filtering, modeling, and output generation.
These parameters provide flexibility on the specific analysis to conduct.

Key parameters include:

- percent_missing: threshold used for missingness-based filtering
- tests: main condition contrasts to be evaluated
- tests_interaction: interaction contrasts or coefficients (optional)
- formula: limma design formula defining main effects and interactions
- reference_condition: reference level for the condition factor
- paired / block_effect: options for paired designs and correlation-aware
    models
- ldv_source: strategy for LDV (left-censored) imputation
- plots: whether to generate output tables and PDF plots

These parameters are illustrated in the workflow section below, but the full
analysis is not executed during vignette rendering.


# Example 1: Workflow with interaction testing (condition-by-batch)

```{r}
# Parameters
experiment <- "HaCaT"
percent_missing <- 50

formula <- formula(~ 0 + condition + batch + condition:batch)
tests <- c(
  "IL13_vs_NoTreated",
  "IL22_vs_NoTreated",
  "COMBO_vs_NoTreated"
)
tests_interaction <- c(
  "IL13.batchday2",
  "IL22.batchday2",
  "COMBO.batchday2"
)
reference_condition <- "NoTreated"

# Workflow
res <- run_proteomics_pipeline(
  path_pgmatrix        = path_pgmatrix,
  path_annotation      = path_annotation,
  path_output          = path_output,
  tests                = tests,
  tests_interaction    = tests_interaction,
  formula              = formula,
  reference_condition  = reference_condition,
  percent_missing      = percent_missing,
  ldv_source           = "per-condition",
  experiment           = experiment,
  plots                = FALSE # Use TRUE to generate outputs
)
```

## Inspecting filtering and imputation outputs (xlsx files)

```{r}
## Inspect filtering and imputation summary tables
library(readxl)

path_filtered <- system.file(
  "extdata",
  "HaCaT_filtered_out_proteins.xlsx",
  package = "ProteinBatcher"
)

path_imputed <- system.file(
  "extdata",
  "HaCaT_Imputation_step.xlsx",
  package = "ProteinBatcher"
)

filtered_df <- readxl::read_xlsx(path_filtered)
imputed_df  <- readxl::read_xlsx(path_imputed)

head(filtered_df)
head(imputed_df)
```

These Excel files are generated by the workflow to make the preprocessing steps
fully transparent.

HaCaT_filtered_out_proteins.xlsx lists proteins removed during the
missingness-based filtering step. This allows users to verify which features
were excluded and why (e.g. insufficient detection across conditions).

HaCaT_Imputation_step.xlsx summarizes the imputation strategy applied to each
protein and sample, distinguishing mean-imputed values from LDV
(left‑censored) imputations.

Together, these tables allow direct comparison of the dataset before and after
filtering and imputation, which is critical for quality control and
interpretability in label‑free proteomics analyses.


Note: For package size reasons, this vignette uses a reduced example of the
imputation summary table. The full table is generated during a standard
workflow run.

## Differential effects - IL13 vs NoTreated contrast

From this point onwards, we focus exclusively on the IL13 vs NoTreated
contrast to illustrate how different effect types are interpreted.

### Main effect (all proteins)

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}
knitr::include_graphics(
  system.file(
    "extdata",
    "HaCaT_MainEffect_AllProteins_IL13_vs_NoTreated.png",
    package = "ProteinBatcher"
  )
)
```

The Main Effect (All Proteins) volcano plot shows the global IL13 effect in 
HaCaT cell lines, ignoring any interaction terms (batch). Every tested protein
is included, regardless of whether its response depends on batch.

How to read it:

- The x‑axis represents the estimated log2 fold change (IL13 vs NoTreated).
- The y‑axis represents statistical significance (adjusted p‑value).

This plot answers the question: “Which proteins are affected by IL13 on average?”

This view is useful for an initial overview, but it does not distinguish stable
effects from batch‑dependent ones.

### Common effect (batch‑independent proteins)

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}
knitr::include_graphics(
  system.file(
    "extdata",
    "HaCaT_CommonEffect_IL13_vs_NoTreated.png",
    package = "ProteinBatcher"
  )
)
```

The Common Effect plot shows proteins whose IL13 response is not significantly
modulated by batch. These proteins pass the interaction test and are therefore
considered stable across batch levels.

How to read it:

- Proteins here show a consistent IL13 effect independent of interaction.
- These are typically the most robust candidates for biological interpretation,
as their direction and magnitude are reproducible across experimental batches

Conceptually, this plot answers:
“Which IL13‑responsive proteins behave consistently across batches?”


### Interaction effect (batch‑dependent proteins)

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}
knitr::include_graphics(
  system.file(
    "extdata",
    "HaCaT_InteractionEffect_IL13.batchday2.png",
    package = "ProteinBatcher"
  )
)
```

The Interaction Effect plot highlights proteins whose IL13 response differs
between batch levels (here: day1 vs day2).

How to read it:

- Significance is driven by the interaction term, not by the average IL13 effect.
- A protein may appear weak or even non-significant in the main effect plot but
still show a strong interaction if its response reverses or changes magnitude
between batches.

This plot answers:
“Which proteins show batch-specific IL13 effects?”

### Deregulogram: interpreting batch-dependent effects

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="80%"}
knitr::include_graphics(
  system.file(
    "extdata",
    "HaCaT_Deregulogram_IL13_vs_NoTreated_batch.png",
    package = "ProteinBatcher"
  )
)
```

The deregulogram compares full IL13 effects between the two batch levels.

- y-axis: full IL13 effect in the reference batch (e.g. day1)
- x-axis: full IL13 effect in the other batch (day2), computed as
main effect + interaction effect

How to read it:

- Proteins along the diagonal behave similarly in both batches.
- Proteins deviating from the diagonal show batch-dependent regulation.
- Highlighted points represent proteins for which the interaction is both
statistically significant and biologically relevant (effect size threshold).

**Key interpretation note**:
The deregulogram is intentionally more restrictive than the interaction volcano.
It is designed to emphasize interpretable, direction-changing or magnitude-shifting
effects, rather than listing all statistically significant interactions. As a result,
the set of highlighted proteins is not expected to match one-to-one with the
interaction volcano, but be more stringent.

# Example 2: Workflow without interaction testing

In some experiments, the main goal is to quantify a global condition effect
without explicitly testing whether that effect depends on a secondary factor.
Typical scenarios include:

- experiments with a single experimental factor and no obvious batch structure,
designs where batches are already well balanced and considered negligible,
analyses where interaction testing would be underpowered or biologically
uninformative.

- In these cases, interaction terms can be omitted from the model and only the
main condition effect is evaluated.

At the level of the workflow, this corresponds to:

- using a design formula without interaction terms, and 
setting tests_interaction = "NA".

Here we use a simple condition-only model:

```{r}
# Workflow parameters (no interaction testing)
experiment <- "HaCaT"
percent_missing <- 50

formula <- ~ 0 + condition
tests <- c(
  "IL13_vs_NoTreated",
  "IL22_vs_NoTreated",
  "COMBO_vs_NoTreated"
)
tests_interaction <- "NA"
reference_condition <- "NoTreated"
```

Running the pipeline

```{r}
res <- run_proteomics_pipeline(
  path_pgmatrix        = path_pgmatrix,
  path_annotation      = path_annotation,
  path_output          = path_output,
  tests                = tests,
  tests_interaction    = tests_interaction,
  formula              = formula,
  reference_condition  = reference_condition,
  percent_missing      = percent_missing,
  ldv_source           = "per-condition",
  experiment           = experiment,
  plots                = FALSE
)

```

When tests_interaction = "NA":

- only main effects are estimated and reported,
- all proteins are treated as responding homogeneously across samples,
- no separation into common versus interaction effects is performed.

This configuration answers the question:

“Which proteins are differentially abundant between conditions on average?”

and is appropriate when there is no strong prior reason to expect
factor-dependent responses or when the analysis is intended as a first check,
global comparison.

# sessionInfo()

```{r}
sessionInfo()
```

