---
title: "ProteinBatcher workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProteinBatcher workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
Suggests:
    knitr,
    rmarkdown,
    htmltools
---

# Overview
Quantitative proteomics is well supported in Bioconductor for individual
downstream tasks such as differential testing (e.g. via limma), missing-value
imputation, or visualization. However, fully integrated downstream workflows
that combine these steps in a single, reproducible, and design-aware pipeline
are comparatively scarce, particularly for label-free proteomics. 

ProteinBatcher addresses this gap by providing a cohesive downstream analysis
framework that focuses on three aspects that are often difficult to combine in
practice:

1. Flexible experimental designs
    - The package is built around limma and supports:
    - multi-factor designs (e.g. condition + batch, condition + sex),
    - interaction testing, 
    - paired and repeated-measures designs,
    - correlation-aware models via duplicateCorrelation.

    This flexibility allows users to model common proteomics scenarios such as
    batch effects, donor effects, or preparation-day effects without rewriting
    analysis code for each experiment.
    
2. End-to-end standardization of downstream steps. ProteinBatcher integrates,
in a single workflow:
    - missingness-based filtering,
    - mean and left-censored (LDV) imputation strategies,
    - differential abundance testing,
    - systematic separation of main, common, and interaction effects,
    - generation of plot-ready result objects.

    By enforcing a consistent structure for inputs, models, and outputs, the
    package reduces ad hoc analysis choices and improves reproducibility across
    experiments.
    
3. Interpretation-oriented visualizations. In addition to standard volcano
plots, ProteinBatcher implements:
    - deregulograms, which compare full effects across the two levels of an
        interaction factor (e.g. batch day 1 vs day 2, female vs male).
        Deregulograms complement interaction volcano plots by emphasizing effect
        concordance and magnitude, rather than statistical significance alone,
        and facilitate biological interpretation of interaction effects.

Importantly, ProteinBatcher is not intended to replace existing Bioconductor
infrastructure, but rather to orchestrate and standardize well-established
methods (notably limma) into a coherent downstream proteomics workflow.
The package is therefore particularly suited for users who repeatedly analyze
proteomics experiments with similar structure but varying designs, and who need
a balance between automation and modeling flexibility.

This vignette documents the workflow and outputs, but does not execute the
full analysis. All examples rely on precomputed results stored in
inst/extdata, ensuring portability and fast rendering during package checks.

# Relationship with the Bioconductor ecosystem

ProteinBatcher is designed to operate within the standard Bioconductor
infrastructure, rather than introducing new data containers or statistical
methods. Its functionality is built by extending existing Bioconductor
components in a consistent downstream workflow.

All intermediate and final results in ProteinBatcher are stored in
SummarizedExperiment objects. Statistical testing in ProteinBatcher is performed
using limma, a well-established Bioconductor package for linear modeling with
empirical Bayes moderation. Missing-value filtering and imputation are
implemented as modular steps that operate directly on SummarizedExperiment
objects. 

# Input data and workflow parameters

ProteinBatcher needs two input data objects and a set of workflow parameters.
This distinction is important: the inputs define the data to be analysed,
whereas the parameters control how the analysis is performed.

## Input data

ProteinBatcher expects exactly two input datasets:

1. Quantitative matrix: A feature-by-sample matrix containing quantitative
proteomics measurements (typically protein groups). Rows correspond to
quantified features and columns to samples. The matrix is assumed to originate
from an upstream workflow (e.g. FragPipe or DIANN), but ProteinBatcher does not
depend on a specific quantification tool.

2. Sample annotation table: A tab-delimited table describing the experimental
design and linking each sample to its biological and technical metadata.

Both inputs are provided in this vignette as example datasets shipped with the
package.

```{r}
# Load package
library(ProteinBatcher)
library(readr)

# Input files shipped with the package
path_annotation <- system.file(
  "extdata", "annotation_HaCaT.tsv",
  package = "ProteinBatcher"
)

path_pgmatrix <- system.file(
  "extdata",
  "2024MK017_HaCaT_Stimulation_1to24_Astral_report.pg_matrix.tsv",
  package = "ProteinBatcher"
)

# Output directory (portable, not used during vignette build)
path_output <- tempdir()

# Inspect annotation structure
ann <- readr::read_tsv(path_annotation)
```

## Sample annotation structure
The annotation table must contain at least a condition column defining the main
experimental comparison. Additional columns enable more complex designs such as
batch correction, paired analyses, and interaction testing.

The following columns are supported:

- file: raw data file identifier
- sample / sample_name: sample identifiers
- condition: main experimental condition of interest
- replicate: technical replicate identifier (repeated measurements of the
    same sample under the same condition)
- block: biological replicate or donor identifier
- batch: batch or preparation factor (e.g. factor to control for against
    condition)

Only columns explicitly referenced in the design formula are required.

## Workflow parameters

In addition to the input data, ProteinBatcher requires several parameters that
define how the analysis is carried out. These parameters do not represent
additional inputs, but control filtering, modeling, and output generation.
These parameters provide flexibility on the specific analysis to conduct.

Key parameters include:

- percent_missing: threshold used for missingness-based filtering
- tests: main condition contrasts to be evaluated
- tests_interaction: interaction contrasts or coefficients (optional)
- formula: limma design formula defining main effects and interactions
- reference_condition: reference level for the condition factor
- paired / block_effect: options for paired designs and correlation-aware
    models
- ldv_source: strategy for LDV (left-censored) imputation
- plots: whether to generate output tables and PDF plots

These parameters are illustrated in the workflow section below, but the full
analysis is not executed during vignette rendering.


# Example 1: Workflow with interaction testing (condition-by-batch)

```{r}
# Parameters
experiment <- "HaCaT"
percent_missing <- 50

formula <- formula(~ 0 + condition + batch + condition:batch)
tests <- c(
  "IL13_vs_NoTreated",
  "IL22_vs_NoTreated",
  "COMBO_vs_NoTreated"
)
tests_interaction <- c(
  "IL13.batchday2",
  "IL22.batchday2",
  "COMBO.batchday2"
)
reference_condition <- "NoTreated"

# Workflow
res <- run_proteomics_pipeline(
  path_pgmatrix        = path_pgmatrix,
  path_annotation      = path_annotation,
  path_output          = path_output,
  tests                = tests,
  tests_interaction    = tests_interaction,
  formula              = formula,
  reference_condition  = reference_condition,
  percent_missing      = percent_missing,
  ldv_source           = "per-condition",
  experiment           = experiment,
  plots                = FALSE # Use TRUE to generate outputs
)
```

## PDF outputs

```{r}
# List available PDFs
pdf_dir <- system.file("extdata", package = "ProteinBatcher")
list.files(pdf_dir, pattern = "\\.pdf$", full.names = FALSE)
```
```{r}
# Embed a PDF (HTML vignette)
library(htmltools)

embed_pdf <- function(fname, height = 600) {
  f <- system.file("extdata", fname, package = "ProteinBatcher")
  tagList(
    tags$embed(
      src = f, type = "application/pdf",
      width = "100%", height = height
    ),
    tags$p(tags$a(href = f, "Download PDF"))
  )
}

embed_pdf("2024MK017_ELMO_001to024_6_HaCaT_Stimulation_Volcano_NoTreated_vs_IL13.pdf")
embed_pdf("2024MK017_ELMO_001to024_6_HaCaT_Stimulation_Volcano_NoTreated_vs_IL22.pdf")
embed_pdf("2024MK017_ELMO_001to024_6_HaCaT_Stimulation_Volcano_NoTreated_vs_COMBO.pdf")
```

# Example 2: Workflow without interaction testing




# sessionInfo()

```{r}
sessionInfo()
```

